String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-04 2023-1-19/sections/algorithms.tex
from recl to cefa

Let $\varphi$ be an RECL constraint. We shall reduce the satisfiability of $\varphi$ to an instance of $\cefadec$.

@@

Let $e$ be a register-representable regular expression over $\Sigma$. We are going to construct a CEFA out of $e$, denoted by $\aut_e$, by the following procedure. 
\begin{enumerate}
\item For each maximal sub-expression $e'$ of $e$ such that $e'$ contains no occurrences of counting operators, an NFA $\aut_{e'}$ can be constructed by a structural induction on the syntax of $e'$. 
%
\item For each sub-expression $(e')^{\{m,n\}}$ with $m \le n$ (resp. $(e')^{\{m, \infty\}}$) of $e$, we construct a CEFA $\aut_{(e')^{\{m,n\}}}$ (resp. $\aut_{(e')^{\{m, \infty\}}}$) as follows. Let $\aut_{e'} = (Q, \Sigma, \delta, I, F)$. From the construction for Kleene star in Section~\ref{sec:pre}, $(\aut_{e'})^* = (Q', \Sigma, \delta', I', F')$, where $Q' = Q \cup \{q_0\}$ with $q_0 \not \in Q$, $I' = \{q_0\}$, $F'= F \cup \{q_0\}$, $\delta' = \delta  \cup \{(q_0, a, q')  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\} \cup \{(q, a, q') \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}$. Then $\aut_{(e')^{\{m,n\}}} = ((r'), Q', \Sigma, \delta'', I', F', \alpha')$, where $r'$ is a register, \\
%
$$
\begin{array}{l c l}
\delta'' & = & \{(q, a, q', (0)) \mid (q, a, q') \in \delta\}\  \cup \\
& & \{(q_0, a, q', (1))  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}\ \cup \\
& & \{(q, a, q', (1)) \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\},
\end{array}
$$ 
%
moreover, $\alpha' = m \le r' \le n$ if $I \cap F = \emptyset$, and $\alpha' = r' \le n$ otherwise.  (Intuitively, if $\varepsilon$ is accepted by $\aut_{e'}$, then the value of $r'$ can be less than $m$.) Finally, $\aut_{(e')^{\{m, \infty\}}}$ is constructed by adapting $\alpha'$ in $\aut_{(e')^{\{m,n\}}}$ as follows: $\alpha' = m \le r'$ if $I \cap F = \emptyset$ and $\alpha' = \ltrue$ otherwise.

%
\item For each sub-expression $e'$ of $e$ such that $e'$ contains occurrences of counting operators but $e'$ itself is not of the form $(e'_1)^{\{m,n\}}$ or $(e'_1)^{\{m,\infty\}}$, from the assumption that $e$ is register-representable, we know that $e'$ is of the form $e'_1 \concat e'_2$, $e'_1 + e'_2$, $e'_1 \cap e'_2$, or $(e'_1)$. For $e' = (e'_1)$, we have $\aut_{e'} = \aut_{e'_1}$. For $e' = e'_1 \concat e'_2$, $e' = e'_1 + e'_2$, or $e' = e'_1 \cap e'_2$, suppose that CEFA $\aut_{e'_1}$ and $\aut_{e'_2}$ have been constructed. Without loss of generality,  we assume that $R_{\aut_{e'_1}} \cap R_{\aut_{e'_2}} = \emptyset$. Then we have $\aut_{e'} = \aut_{e'_1} \concat \aut_{e'_2}$, $\aut_{e'} = \aut_{e'_1} \cup \aut_{e'_2}$, $\aut_{e'} = \aut_{e'_1} \cap \aut_{e'_2}$ respectively. 
\end{enumerate}

@@

@@

The most important point of our approach is encoding counting operators by registers rather than explicitly unfolding them. In this section, we discuss the encoding in detail. A counting operator is called \emph{handled} if it is not the sub-regex of complement, closure, or another counting, and called \emph{unhandled} otherwise. For example, $(\Sigma \setminus a)^{\{m,n\}}$ is handled, while $((\Sigma \setminus a)^{\{m,n\}})^*$ is unhandled. We first discuss the encoding of handled counting operators and then the encoding of unhandled counting operators.

@@

%   Given two CEFAs $\aut_1 = (Q_1, \Sigma, \delta_1, q_{I_1}, F_1, R_1, \theta_1)$ and $\aut_2 = (Q_2, \Sigma, \delta_2, q_{I_2}, F_2, R_2, \theta_2)$ with $R_1\cap R_2 = \emptyset$, the product of them, denoted by $\aut_1 \times \aut_2$, is defined as $(Q_1\times Q_2, \Sigma, \delta, (q_{I_1}, q_{I_2}), F_1\times F_2, R_1\cup R_2, \theta)$ where
%   \begin{itemize}
%     \item $\delta$ comprises the tuples $((q_1, q_2), a, (q_1', q_2'), \myvec{v_1}\cdot \myvec{v_2})$ such that $(q_1, a, q_1', \myvec{v_1})\in \delta_1$ and $(q_2, a, q_2', \myvec{v_2})\in \delta_2$.
%     \item $\theta((q_1, q_2))$ is $\theta_1(q_1)\wedge \theta_2(q_2)$ for each accepting state $(q_1, q_2)\in F_1\times F_2$.
%   \end{itemize}
% \end{definition}
% With the help of product, we can combine the CEFAs of regex-counting and string-length. For example, the CEFA of $x$ in the string constraints $x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c \wedge |x| > 120$ is the product of the CEFA of regex $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$, the CEFA of regex $\Sigma^*$ and the CEFA of length constraints $|x| > 120$.

>>>>

Let $\varphi$ be an RECL constraint and $x_1, \cdots, x_n$ be an enumeration of the string variables occurring in $\varphi$. Moreover, let $\varphi \equiv \varphi_1 \wedge \varphi_2$ such that $\varphi_1$ is a conjunction of regular membership constraints of $\varphi$, and $\varphi_2$ is a conjunction of length constraints of $\varphi$.
We shall reduce the satisfiability of $\varphi$ to an instance of $\cefadec$.

@@

Let $e$ be a register-representable regular expression over $\Sigma$. We are going to construct a CEFA out of $e$, denoted by $\aut_e$, by the following procedure. 
\begin{enumerate}
\item For each maximal sub-expression $e'$ of $e$ such that $e'$ contains no occurrences of counting operators, an NFA $\aut_{e'}$ can be constructed by a structural induction on the syntax of $e'$. 
%
\item For each sub-expression $(e')^{\{m,n\}}$ with $m \le n$ (resp. $(e')^{\{m, \infty\}}$) of $e$, we construct a CEFA $\aut_{(e')^{\{m,n\}}}$ (resp. $\aut_{(e')^{\{m, \infty\}}}$) as follows. Let $\aut_{e'} = (Q, \Sigma, \delta, I, F)$. From the construction for Kleene star in Section~\ref{sec:pre}, $(\aut_{e'})^* = (Q', \Sigma, \delta', I', F')$, where $Q' = Q \cup \{q_0\}$ with $q_0 \not \in Q$, $I' = \{q_0\}$, $F'= F \cup \{q_0\}$, $\delta' = \delta  \cup \{(q_0, a, q')  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\} \cup \{(q, a, q') \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}$. Then $\aut_{(e')^{\{m,n\}}} = ((r'), Q', \Sigma, \delta'', I', F', \alpha')$, where $r'$ is a register, 
%
$$
\begin{array}{l c l}
\delta'' & = & \{(q, a, q', (0)) \mid (q, a, q') \in \delta\}\  \cup \\
& & \{(q_0, a, q', (1))  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}\ \cup \\
& & \{(q, a, q', (1)) \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\},
\end{array}
$$ 
%
moreover, $\alpha' = m \le r' \le n$ if $I \cap F = \emptyset$, and $\alpha' = r' \le n$ otherwise.  (Intuitively, if $\varepsilon$ is accepted by $\aut_{e'}$, then the value of $r'$ can be less than $m$.) Finally, $\aut_{(e')^{\{m, \infty\}}}$ is constructed by adapting $\alpha'$ in $\aut_{(e')^{\{m,n\}}}$ as follows: $\alpha' = m \le r'$ if $I \cap F = \emptyset$ and $\alpha' = \ltrue$ otherwise.

%
\item For each sub-expression $e'$ of $e$ such that $e'$ contains occurrences of counting operators but $e'$ itself is not of the form $(e'_1)^{\{m,n\}}$ or $(e'_1)^{\{m,\infty\}}$, from the assumption that $e$ is register-representable, we know that $e'$ is of the form $e'_1 \concat e'_2$, $e'_1 + e'_2$, $e'_1 \cap e'_2$, or $(e'_1)$. For $e' = (e'_1)$, we have $\aut_{e'} = \aut_{e'_1}$. For $e' = e'_1 \concat e'_2$, $e' = e'_1 + e'_2$, or $e' = e'_1 \cap e'_2$, suppose that CEFA $\aut_{e'_1}$ and $\aut_{e'_2}$ have been constructed. Without loss of generality,  we assume that $R_{\aut_{e'_1}} \cap R_{\aut_{e'_2}} = \emptyset$. Then we have $\aut_{e'} = \aut_{e'_1} \concat \aut_{e'_2}$, $\aut_{e'} = \aut_{e'_1} \cup \aut_{e'_2}$, and $\aut_{e'} = \aut_{e'_1} \cap \aut_{e'_2}$ respectively. 
\end{enumerate}

@@

For non-register-representable regular expressions, we first transform them into register-representable regular expressions by unfolding all the non-register-representable occurrences of  counting operators, then utilize the aforementioned procedure to construct CEFA. For instance, $(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ is transformed into $(aa(\varepsilon + a + aa + aaa + aaaa) \concat b^*) \setminus (a+b)(a+b)(a+b)(a+b)(a+b)^*$.

For each $i \in [n]$, let $x_i \in e_{i, 1}, \cdots, x_i \in e_{i, l_i}$ be an enumeration of the regular membership constraints for $x_i$ in $\varphi_1$.  Then we can construct CEFA $\aut_{i, j}$ from $e_{i, j}$ for each $i \in [n]$ and $j \in [l_i]$. Moreover, we construct another CEFA $\aut_{i, 0}$ for each $i \in [n]$ to model the length of $x_i$. Specifically, $\aut_{i,0}$ is constructed as $((r_{i,0}), \{q_{i,0}\}, \Sigma, \delta_{i,0}, \{q_{i,0}\}, \{q_{i,0}\}, \ltrue)$, where $r_{i,0}$ is a fresh register (in particular, $r_{i,0}$ does not occur in $\aut_{i,1}, \cdots, \aut_{i, l_i}$) and $\delta_{i,0} = \{(q_{i,0}, a, q_{i,0}, (1)) \mid a \in \Sigma\}$. 
Let $\Lambda_{x_i} = \{\aut_{i,0}, \aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for each $i \in [n]$. Moreover, let $\varphi'_2 \equiv \varphi_2[r_{1,0}/|x_1|, \cdots, r_{n,0}/|x_n|]$.
Then the satisfiability of $\varphi$ is reduced to the nonemptiness of CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi'_2$.

@@

%%%%%%%%%%% original texts by denghang %%%%%
%%%%%%%%%%% original texts by denghang %%%%%
\hide{
The most important point of our approach is encoding counting operators by registers rather than explicitly unfolding them. In this section, we discuss the encoding in detail. A counting operator is called \emph{handled} if it is not the sub-regex of complement, closure, or another counting, and called \emph{unhandled} otherwise. For example, $(\Sigma \setminus a)^{\{m,n\}}$ is handled, while $((\Sigma \setminus a)^{\{m,n\}})^*$ is unhandled. We first discuss the encoding of handled counting operators and then the encoding of unhandled counting operators.

@@

%   Given two CEFAs $\aut_1 = (Q_1, \Sigma, \delta_1, q_{I_1}, F_1, R_1, \theta_1)$ and $\aut_2 = (Q_2, \Sigma, \delta_2, q_{I_2}, F_2, R_2, \theta_2)$ with $R_1\cap R_2 = \emptyset$, the product of them, denoted by $\aut_1 \times \aut_2$, is defined as $(Q_1\times Q_2, \Sigma, \delta, (q_{I_1}, q_{I_2}), F_1\times F_2, R_1\cup R_2, \theta)$ where
%   \begin{itemize}
%     \item $\delta$ comprises the tuples $((q_1, q_2), a, (q_1', q_2'), \myvec{v_1}\cdot \myvec{v_2})$ such that $(q_1, a, q_1', \myvec{v_1})\in \delta_1$ and $(q_2, a, q_2', \myvec{v_2})\in \delta_2$.
%     \item $\theta((q_1, q_2))$ is $\theta_1(q_1)\wedge \theta_2(q_2)$ for each accepting state $(q_1, q_2)\in F_1\times F_2$.
%   \end{itemize}
% \end{definition}
% With the help of product, we can combine the CEFAs of regex-counting and string-length. For example, the CEFA of $x$ in the string constraints $x \in (\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}} \wedge x \in \Sigma^* c \wedge |x| > 120$ is the product of the CEFA of regex $(\Sigma \setminus a)^{\{5, 60\}} (\Sigma \setminus b)^{\{5, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$, the CEFA of regex $\Sigma^*$ and the CEFA of length constraints $|x| > 120$.
}
%%%%%%%%%%% original texts by denghang %%%%%
%%%%%%%%%%% original texts by denghang %%%%%

### 2023-05-04 2023-1-19/sections/preliminaries.tex
from recl to cefa

We will also use standard quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulae, which are typically ranged over by $\phi, \varphi$, etc. 
For a set $X$ of variables, we use $\Phi(X)$ to denote the set of quantifier-free LIA formulas whose free variables are from $X$. 
For convenience, we use $\varphi(\vec{x})$ with $\myvec{x} = (x_1, \cdots, x_k)$ to denote an LIA formula $\varphi$ whose free variables are from $\{x_1, \cdots, x_k\}$. For an LIA formula $\varphi(\vec{x})$ with $\myvec{x} = (x_1, \cdots, x_k)$ and $\myvec{n} = (n_1, \cdots, n_k) \in \Int^k$, we use $\varphi[\myvec{n}/\myvec{x}]$ to denote the formula obtained by replacing $x_i$ with $n_i$ for each $i \in [k]$.
%For convenience, we usually assume that a set $X$ of variables is a vector. 
%, which is essentially Presburger

>>>>

We will also use standard quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulae, which are typically ranged over by $\phi, \varphi$, etc. 
For a set $X$ of variables, we use $\Phi(X)$ to denote the set of quantifier-free LIA formulas whose free variables are from $X$. 
For convenience, we use $\varphi(\vec{x})$ with $\myvec{x} = (x_1, \cdots, x_k)$ to denote an LIA formula $\varphi$ whose free variables are from $\{x_1, \cdots, x_k\}$. For an LIA formula $\varphi(\vec{x})$ with $\myvec{x} = (x_1, \cdots, x_k)$ and $\myvec{t} = (t_1, \cdots, t_k)$ be a tuple of integer terms, we use $\varphi[\myvec{t}/\myvec{x}]$ to denote the formula obtained by replacing (simultaneously) $x_i$ with $t_i$ for every $i \in [k]$.
%For convenience, we usually assume that a set $X$ of variables is a vector. 
%, which is essentially Presburger
